Long ago the old men used to tell about various things. They would tell the young men about different things by way of instruction. They would spend two or three days going about to different homes (telling stories). They would say, ``This is the way we used to live, the way we used to go about, the way we used to gain our livelihood,'' instructing the youths. And on the women's side, they in turn talked to young ladies.

In this connection my grandmother would get me up very early in the morning to make a long run, telling me that I would that way become a strong, husky man.

Once when I was herding sheep I came back to the hogan to find my grandfather, one called Man Who Lassoes, sitting there. He used to spend most of his time in the area around Adobe Sticks Up. The old man called Man Who Lassoes was quite a character.

He was still there when I returned. So a sheep was immediately caught for him and butchered. Then the fattest ribs were put in the fire (to roast) for him. But for me -- only the entrails were roasted for me. So I made a fuss about that and started to make trouble.
``How come he gets the choice meat when he does none of the herding? I'm the one who should have the ribs to eat. He's the one who ought to be eating the entrails,'' I made a fuss saying.

Then I sulked around and didn't eat at all. And after he had really eaten his fill he went off to one side and lay down on his back.

After sundown, when it became really dark, I went back inside to the place where I would sleep. But my grandmother got me back up.

``Sit here, my son. Your grandfather is going to tell us about something. Listen to him carefully," she told me.

``Not too awfully long ago, over there at Red Rock, the Agent was thrown out. And a man by the name of Black Horse caused the trouble. This same man used to go by the name of Butcher Squeezed Together In the Middle,'' said my grandfather as he began his story. He told it as though it were something that had just occurred. Here's how this story went:

``I was resting in the shade at Lizard Spring when a horseman came up. He said, `Do you know what? Over there at Red Rock they've tossed out the Agent. The man called Black Horse, or Butcher Squeezed Together In the Middle, is causing the trouble. That's what he announced to me,' said my grandfather.

``My horse was grazing nearby, so I caught and saddled him. Then I put my double barreled gun into its saddle scabbard and strapped on my pistol. I also strapped on my small pistol. I also tied on the rear a package containing an ample supply of ammunition. Then I went down into the Flat Rock Valley.

``I met up with a number of husky men and asked them if any would like to go with me. I told them that a man known as Black Horse was causing trouble over there, but they all refused. I kept on going nevertheless. Suddenly I came upon a man of the Red Streak Into Water Clan called Little Policeman [from Chinle]. He was a strong, husky young fellow (he is now an old man),'' said my grandfather.

``'I hear that the man called Black Horse is causing trouble. Maybe you'd like to join up with me and go over there. I hear they've tossed out the Agent. It's because I got word of it that I'm going', I told him,'' said my grandfather.

``'Why not? Sure, I'll join you. We'll go,''' he said as he strapped on his gun. Then we started out together.

``We went down into the Canyon De Chelly and came up out of there at a place called Sparse Group Of Pines Extends Up Out. From there we went to Lukachukai and up onto the mountain. Then we went onward to the rim of the mountain to where the trail descended into Cove. At that time it was only a horse trail, There, at that time, there was a fence of boughs with a pole gate through which passed a trail," said grandfather.

``[He] dismounted to open the gate, [the one called Little Policeman]. There was danger lurking on every side. It was potentially a perilous area. Just as he got down from his horse and took hold of a pole to remove it, a cottontail jumped up right at his feet. Little Policeman was so taken back that he nearly fell over right there.

``We resumed our journey and came to a high point that overlooked the surrounding country, and out there in front of us there were people engaged in training maneuvers. They were Navajos. They would come dashing out lying close beside their horses in such a way that there appeared to be no riders. And their maneuvers took many other forms. On the high points they had lookouts stationed, but somehow we passed through without being sighted. that's probably because they were on lookout. Somehow we crossed their lookout points without them seeing us,'' [he said.]

``We went on until we came to Black Horse's hogan. There was quite a gathering there. When we arrived we heard Black Horse talking inside a large hogan, telling how they would do and how they would win.

``Then he was informed of our coming. `Who the devil comes without saying a word? Tell them to come in,' he said. So we went in.

{``}`You who are called Black Horse -- you who go by the name of Butcher Squeezed Together In The Middle -- you think you're famous, the way people tell about you. Come on and see if you can swallow me head first,' I said to him,'' said my grandfather [Man Who Lassoes].

{``}`Oh, oh you've got me,' said Black Horse.''

``And then all the people who were gathered there, and to whom he had been talking, roared out at him, saying, ``We knew that's what you would say. We knew it all the time. You went ahead, even though you're a coward. When even a fellow Navajo spoke to you, you said, `You've got me!'{''}


``And then I really told Black Horse off, but good," said my grandfather.

``This trouble that you're making is uncalled for. Over toward Tuba City a fellow known as Pollen caused trouble and got nothing for it. Over by Aneth a man called He Who Has Supernatural Power caused trouble. And over toward Beautiful Mountain a man by the name of Bizhoshi made a commotion. None of them gained a thing. They merely brought hardship upon themselves. So don't sulk, my baby. It's uncalled for, and it is bad behavior. You will gain nothing by it,'' my grandfather said to me. He told me this story and gave me instructions on top of it. I now remember this well.
