In the Fort Sumner Treaty we were told to place our children in school. So when we got back policemen were sent from Fort Defiance down toward Round Rock to carry out this provision. To get all of the children they went from hogan to hogan. The leader of one of the parties was a man named Charlie. Others in this party were one called Bobbed Hair, one by the name of Bead Clan Gambler, one called Slender Curly Hair another called Slender Silver Maker, and the one who was known as The Interpreter (Chee Dodge), as well as several others. After we volunteered some men, they would travel down there.

For a long time the men and womenfolk had held back their children. So on the dates set for bringing the children to Round Rock, very few were brought in. When the children were taken there a man with the name of Black Horse said, ``No,'' and stood against the children being taken away to school. There was a man by the name of Limper, who may still be hunched with old age (still alive). There was one called Slow One, and there were Canyon De Chelly Man, Tall Bitter Water Clansman, Ugly Knife, Sucker, Weela, and Gray Haired Man. All of these said, ``No,'' and stood against the proposal. And then many young men joined with them. All of these men are now perhaps dead.

Here at this meeting to get children together for school the man known as Black Horse stubbornly balked on it. ``What the devil, you can jerk our children away from us if you want to. If you want trouble over this matter go to it,'' he shouted. Several others who were of like mind were behind him. In fact, they had probably conspired with him.

The one called Little Chief (Shipley) came also. He was the Agent. But when a hot argument got under way in the meeting, the Navajos threw the Agent out. Then things really happened. The man called Bead Clan Gambler got the Agent back from those who had laid hands on him. This Bead Clan Gambler was a husky man. As the mob made away with the Agent he dashed in among them and grabbed the Agent up under one arm (like a football). When the mobsters would have jumped him he straight-armed them and bowled them over backward, and running hard he beat them to a flour storage room. When he had gotten the Agent in there he piled things against the door from within, and the mob was frustrated. The Interpreter (Chee Dodge) also took refuge in the flour storage room. It was said that they stayed inside for several days without going outside. And while they were in there they dirtied the flour.

The news of this trouble spread all about, even up into the Monument Valley. ``They're not going to take our children away, that's all there is to it.'' People were ready to fight.

Some said that no soldiers had come to the scene of the trouble. Some said that bear hunters had been brought to the rescue, but the truth is that it wasn't these. That was just a tale. Soldiers came from Fort Defiance. They were sent from there, and those who came moved up on top of Row of Willows (near Wheatfields). At that time there was a beautiful meadow there. There was not a single wash at that time. There they were encamped. When they moved again they went to the burned trading post at Black Rock Spring (near Tsailee). After camping there for two days they moved on. They moved down to Bue Clay Point (near Round Rock).

This was not the only time a school party went out. They went out many times. Summer and winter they would go about telling the people to place their children in school.

When the soldiers who were sent for arrived to fight, my older sister, who was the wife of Weela, acted as a peacemaker. Here she acted as a go-between to restore the peace. One man took credit for restoring peace. He was a man called White Man. I heard him saying that he was the one who restored the peace. [...] ``You are a liar. You just made that story up. It was my sister who restored the peace. I know it. I am sure of it,''[ I told him].́ So that is how Black Horse made trouble by holding back the children.

Now school is a wonderful thing. They had no reason for keeping us out of school, I'm convinced of that. School is not a thing of no value. It is something to be longed for and sought after. Had I gone to school I wonder how I would be today. By school we mean an endless learning. It's a means for accomplishment without end.

People who have children, and who put them in school, are right in so doing. I think they are indeed lucky children. I was one of those who spoke in favor of schools in days gone by. Not long ago a paper was brought up before us in a meeting. They said, ``Here's a list of the men who asked for education, and here's one of the men. These men knew the advantages of schooling.'' They said this, mentioning my part in it. When they mentioned me in this connection I almost wept. I was right when I took a stand. Had I gone to school I often wonder how good a leader I might have become.

Long ago our forbears went after one another with weapons over this question of schooling. On account of differences over education they threw out an Agent. It nearly brought tragedy. When people were about to come to blows my sister restored the peace. These things happened just about the time when my mind had become mature enough to reason and remember. So, I know whereof I speak. And to many things which ocurred I was a witness, while in others I took an active part. So the people nearly came to blows over schools. But now education is, without doubt, the right thing. So go on children. Go to school. Study hard. In the future you will profit by it.