I am a medicine man. I am going to tell the story I got long ago from a man called Man With A Moustache, who was my grandfather. It was twenty-seven years ago that he died of old age, at the age of a hundred and two years; I shall now begin to tell you his story, the one which he himself told to me. It is said that long ago in the Underworld, there existed the mountains known as Sierra Blanca Peak, Mount Taylor, San Francisco Peak, La Plata, Huerfano, and Gobernador Knob. And people were in existence living there. It was then that Coyote did something. Somewhere a river was flowing, it is said. The offspring of Grabs Things in Deep Water (a Water Monster) was floating about when Coyote approached it and tried to get it. He lassoed it with a sunbeam and pulled it ashore. For that reason, things began to go badly for The People. All the birds that fly flocked together for no apparent reason. Something black kept rising up and receding. Cold came from the south, the west, and the north, it is said. The People wondered what it was. They looked and found it to be water.

It was then that First Man and First Woman picked up Sierra Blanca Peak. They picked up Mount Taylor, San Francisco Peak, La Plata Mountain, Huerfano, and Gobernador Knob. They gathered up all the seeds of plants. In a certain place, there was a mountain, and all the creatures climbed to its summit. Then the mountain began to absorb the water and dissolve itself into mud with them on it. At this point, they became desperate and planted a tree to ascend by—a fir tree, they say—but it only grew to the present height of such trees. They were in desperation, and while they were in the treetop, two people (men) of some kind arrived there nearby. ``Who can they be? Someone go and tell them what the situation is,'' someone said. At that point, they were informed, and they came over. ``Could it be that you might do something?'' they were asked. ``That's the way to talk,'' they said, and room was made for them. One of them started off to one side (out of sight). His companion was asked, ``Who is that?'' ``He is my maternal uncle. He is the Sun,'' he replied. He came back in as soon as he felt sure that they had named him. When that one came back in, the other one started away. Then this one was asked, ``Who is that?'' ``He is the Moon. He is my maternal uncle,'' he said. That is why The People do not speak each others' names in each others' presence. Here is where the precedent was established.

This is about the only way, he said, as he stuck his big reed flute into the ground. His companion had a pine tree flute. Before long, there appeared a white speck where the flute extended against the sky. Then they got into it, got into it, got into it -- moved. The Turkey was last. His tail stuck into the foam, so the white tip is the foam, they say. The People climbed up to one of the sky ledges, and there all those who could dig tried in vain. ``It's next to impossible, my grandchildren. There are layers of rock,'' said Gopher. (That is why he doesn't dig straight down; instead, he merely digs a little way down in the top layer of soil and thereafter makes his hole horizontally.) Then the Locust went over, and he got through. He climbed up out on La Plata Mountain at the place called Where They Emerged. They found a lake of water standing there, it is said, impounded by Gray Mountain and Tobacco Mountain. There was a great expanse of water in the midst of which Locust appeared as a black speck. There was something called Those With a White Spot On The Nose (a monster). These could be seen floating leisurely in the four directions. They appeared as black marks. The one on the south was blue; the one on the west was yellow; the one on the north was white, and the one on the east was black. Each of them was holding up an obsidian knife, which was their weapon.

It was then the black thing from the east rushed at him. ``Don't Blink! Don't Blink! Look out!'' said the Wind to him, said the Child of the Wind to him, whispering into his ear. That was one of our Holy Ones who used to tell us things. At this point, the Monster made a pass at him with the weapon. Then the one from the south rushed at him and did the same. Then they initiated him from the four directions (as the Ye'ii initiate little children so the latter will not go blind). ``There are many people from down below. We were living in plenty in the Underworld, but the water was rising on us, and that is why we came up,'' it was said. ``We are to live here,'' The People say. It is thus that it has come about. ``So all of this water should be allowed to flow away,'' said the Locust. ``No,'' he was told. ``No,'' he said, it is said. ``Then I also refuse (to go back down into the Underworld),'' said Locust. This (dialogue) took place four times.

Then he (Locust) pulled out something fletched with tail feathers. There was a passage through his chest[. T]hrough which to pass these [(]feathered[)] objects. [(]He ran the two objects across his mouth[)] and then through the passage in his chest, sticking one in from one side and one in from the other and then crossing his arms to pull them out in opposite directions. ``All right, let's see if you can do what I did. If you can do what I did, then we will let the water stand,'' said Locust. This is what happened. (He used to blink his eyes until the Wind told him not to do so, and that is why Locust does not blink even today.) Then the black things said, ``No,'' that they could not do what he had done. So it was that he won the water.

Mountain Sheep was notified, and with a sheep's horn, he dug the way out for the water. He dug through Tobacco Mountain (Grand Canyon Mountain), and the instrument with which he dug still stands over there where the mountains run together. That's the Sheep's Horn, my grandfather said. It happened like this, it is told. Then Badger came and made the hole from the Underworld larger, and the people emerged through there. Badger really got himself covered with mud, and that is how his belly became black. But the water from the Underworld was still rising behind them, and presently from the hole, there stuck the horn of Grabs Things In Deep Water.

``What's the matter? Why are we being treated in this way?'' The People said. ``While we were living down there in the Underworld, that one called Coyote really got into a lot of trouble and mischief. Could it be because he has the baby of Grabs Things In Deep Water?'' they said. They searched him, and there it lay in his armpit. It fell out, and when this happened, they knew. They suggested throwing it back, but Coyote protested. ``No. What will we have to live by? I did it so we would have this (as a god) to live by,'' Coyote said. There, between the horns of Grabs Things In Deep Water, there was swirling black foam. A kind of Hard Goods called ``perfect'' was placed between the horns. In exchange for that, the lapping water receded, and some of the foam was gathered by The People.

``If this one is thrown back into the water, then what will we live by?'' they said. Grabs Things In Deep Water, the one who had stuck his head up, went back down, and his baby came up above to become the thunder and lightning. That's what we live with when it rains. It is thus, it is said. This earth was just a soft mass, it is said. The wind was notified, and for four days it blew. But the earth hardened and packed only slightly. Some foxtail grass was planted. Those of the squirrel family, such as the Fox Squirrel, brought the seeds of plants and nuts up from below as food. A tree was planted, and it held firm by the roots. First Man took the mountains he had picked up. A song was heard, and he put Sierra Blanca Peak in its proper location over in the east. ``Let this mountain be placed far away; let it lie far away so our thoughts will be long,'' he said.

He set Sierra Blanca Peak in its position, and it was of White Shell. Then Mount Taylor was set in its position, and it was of Turquoise. These two mountains were put in their positions, and both \textit{Haashch'ééłti’í} and \textit{Haashch'éóghan} (two supernatural beings or divinities) took their place in them. Then San Francisco Peak was set in position. It was placed in the west and was of Abalone. \textit{Haashch'ééłti'í} and \textit{Haashch'éóghaan} took their places in it. La Plata Mountain was set up in the north. It was adorned with Jet, and \textit{Haashch'ééłti'í} and \textit{Haashch'éóghaan} took their place in it. There it happened again, it is said. So, in the form of these Sacred Mountains, was our Mother made for us. Sierra Blanca Peak is our Mother. Among the white people, the missionaries speak like that.

This mountain, called Sierra Blanca Peak, is our missionary (i.e., our religion). In accord with it, we live. In the midst of these four Sacred Mountains that were placed, there we live. With that, we who are The People are the heart of the world. These Sacred Mountains that were placed for us are the boundaries of our domain. They are our boundaries. I am a medicine man, so I have some of the soil from these Sacred Mountains. These were established with White Shell, Sheep, Domestic Animals, Maidens, and Youths. In accord with that do we live. Those of The People who possess livestock possess it by virtue of the power derived from the soil of the Sacred Mountains. And we who are medicine men -- this is our way. We have the Mountain Prayers and the Mountain Songs. [Now it is like this.]