In The First Mother-in-law, Tony Tsosie narrates a tale set at Gobernador Knob, where First Man, First Woman, White Shell Woman, and Changing Woman were gathered. At dawn, First Man notices a small cloud hovering over the peak. Each morning, he observes the cloud descending until it obscures the mountain completely. After four days of this mysterious occurrence, First Woman urges him to investigate. When First Man arrives at Gobernador Knob, he hears a noise repeated four times, and finds a baby. Just then, Talking God appears from the east, and claims the child. First Man and Talking God debate who will raise the child until Talking God relents. First Man takes the child back to First Woman, and the baby matures quickly to a young woman and is able to perform errands within four days.

One day, while gathering wood, the young woman is held down by an unseen force. Turning around, she discovers it is the Sun, adorned with turquoise earrings, who instructs her to ask her parents to build a separate brush shelter. That night, she sleeps in the shelter and is visited by the Sun, who lies beside her under a robe. Sun tells the woman to show the robe to her parents for proof of his visitation. 

The story provides several cultural elements, such as the centering of the number four in rep̨eated acts. It also explains the origin for the mother-in-law tradition: Sons-in-law provide merely a gift to their mother-in-laws, but cannot see them, and as such, the mother-in-law is called ``the one whom one does not see.''