There are stories which tell of the events which took place long ago on this earth of ours. They say that this story took place at Gobernador Knob.

Those called First Man and First Woman, and White Shell Woman and Changing Woman had come together and they were sitting. Just as dawn was breaking First Man went outdoors to urinate; he glanced over at the peak of Gobernador Knob, and saw a little cloud hanging there, it is said. When he went back inside he said, ``I wonder why it is that there is a little cloud hanging over there on the summit of Gobernador Knob?'' he said, it is said. ``Why, it's true. I wonder why it is?'' exclaimed First Woman while she ran outside too.

The sun set again. And then with the first streak of dawn it was that way again. When First Man got up he saw that it was that way again. But this time the cloud hung a little lower down on the peak. When he went back inside he again mentioned it. And the Woman again ran out. ``It sure is. I wonder why it is?'' she said again. And it was that way all day.

When dawn came First Man got up. He again looked at Gobernador Peak and the dark cloud was still farther down. Only the base of the mountain was visible. ``What can it be over there at Gobernador Knob?'' It is said that right now there is only a visible section underneath the dark clouds that are covering it,'' he said. ``Let me see,'' again said the Woman. ``It sure is,'' she again said. And it was that way again all day.

And then at the first streak of dawn he got up and again looked at Gobernador Knob. And he found that it had become invisible. This made the fourth time. This being the fourth time, First Woman said to him, ``Go see what you can find out. Investigate it. Why does it do that?'' she said to him.

So he set out for there. When he arrived there he found that visibility was poor. One could not see even along the edge of it. He went some distance, and then decided to turn back. Just as he turned to start back he heard something. But he looked around without seeing anything. He started off again, but again there was a noise. And it happened that way four times. He stopped in his tracks and a corn fly chirped. Then he went back to where the noise came from, and found a baby lying there, it is said. And at that moment he heard Talking God speak over toward the east. ``Woo, woo,'' he said. After he had spoken four times he appeared right there. ``I've come for this baby,'' said Talking God. ``I also came for her,'' First Man told him. For a long time they argued about this matter. ``I will surely grow by means of it. It will be like this in the future,'' they said to each other. ``I will raise her with the pollen from the plants and with beautiful flowers,'' said Talking God. ``So will I. I have everything with which to raise her. I will raise her with the pollen from plants, and with berries,'' said First Man. They argued that way for a long time, and finally the one called Talking God stepped aside. And then he went back into the dawn, it is said.

Then First Man picked up the baby and started toward First Woman with it. With care the baby grew up quickly. Before the fourth day following that on which she was picked up, she could be sent on errands. She could bring fire wood on her back. On the fourth day she again went after wood and Sun came down upon her. On the fourth day she again went after wood and Sun came down upon her. She was trying to get up with the load of wood on her back, but something was holding it down. Something kept holding it down and she could not lift it up. After the fourth try she said, ``What is holding my wood down?'' As she said this she turned her head around and there was Sun covered with (turquoise) earrings. So she knew that it must have been Sun who was holding down her wood. ``Your father and mother know the purpose of my actions. Tell them to build a brush shelter a little way from the hogan,'' Sun said as he went back.

So she picked up the firewood, and when she got back home she told her mother and father. So the brush shelter was made for her. She sat down alone there to spent the night. There was nothing whatsoever inside the shelter. There was not even bedding there. When she became sleepy she just lay down on the ground. She slept for quite a while. At midnight she found that someone was lying beside her. It was the same one who had come to her when she had gone for wood. Right there they slept together. They spent the night beneath a robe.

``Your mother and father will not believe, perhaps. So, give them this robe as proof (so that they believe). The custom of building a separate home in connection with those who marry will go on into the future. It will be the custom,'' said Sun as he went back up. That is the story they tell about the origin of the name ``the one whom one does not see'' (mother-in-law). (Because Sun did not look up on the girl's mother, but merely sent a gift. The groom still gives the mother-in-law a gift at marriage, but doesn't look at her.)