When I had a husband and a lot of sheep I was happy. Then when it came my sheeps' turn to be acted upon, they were all driven away from me. They were driven into a corral and I was told to select the ones that were the best. That hurt me and I just bowed my head. I just sent my children over and they selected some.

We became despondent over it. My husband became sick from worry, and these officials went about in their cars even at night. Early in the morning one would be going about. They even interrupted our sleep. They did this on account of our sheep and our way of life. When we drove our sheep to the dipping vat some more were taken away. It was there that my husband said, ``You people are indeed heartless. You have now killed me. You have cut off my arms. You have cut off my legs. You have taken my head off. There is nothing left for me. This is the end of the trail,'' he said. This happened, and it wasn't long before my husband fell ill. It was no doubt the worry that sickened him. He was sick all winter, and at the beginning of spring he died.

These two events, the loss of my sheep and the loss of my husband, made me feel terribly unhappy. I was despondent, and I am still so today. I had five sons then, and seven daughters. We were all dependent on the small amount of stock granted to us on the permit (lit. we all climbed onto this tiny permit.) We could not always make ends meet (lit. some of us would fall off it and would grab about for something to reestablish our balance and get us back onto it.) In the course of getting one back onto it, another would fall off, so we really had to work closely together in order to stay on this permit. We suffered from everything, from hunger, from lack of meat and from despondency. The Special Grazing Regulation is like a killing disease from which one cannot sleep. We cannot live because of it. One will not walk pleasantly.